---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

The first step to any research is to load the necessary packages and the data 
itself.

```{r}
    library(ggplot2)
    activity <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Next we'll determine our total steps for each day.

```{r}
    total <- tapply(activity$steps, activity$date, sum, na.rm=TRUE)
    total
```

Not bad information, but a histogram would be easier to read.

```{r}
    hist(total)
```

We might like to see the mean and the median as well.

```{r}
    mean(total, na.rm=TRUE)
    median(total, na.rm=TRUE)
```

## What is the average daily activity pattern?

Now lets look at the intervals.  Specifically, the average number of steps taken
per interval across all the days. We also want to put this into a data frame 
that we can then use for our plot.

```{r interval averages}
steps <- tapply(activity$steps, as.factor(activity$interval), mean, na.rm = TRUE)
steps <- as.vector(steps)
intervals <- as.vector(unique(activity$interval))
byint <- data.frame(steps, intervals)
```

Now we can create our time series plot!

```{r time series plot}
plot(byint$intervals, byint$steps, type = "l")
```

Clearly, more steps are taken in the morning, around lunchtime, and in the 
evening. This makes sense with work schedules. But during what interval are the 
most steps taken? Let's find out!

```{r interval with max steps}
byint$intervals[which.max(byint$steps)]
```

So, at 0835 the most steps, on average, were taken each day!

## Imputing missing values

Looking at the original data, there are a lot of NAs. How many, you ask?

```{r number of NAs}
sum(is.na(activity$steps))
```

2304 NAs! That's a lot. Let's fill in all the NAs. We'll use our average steps
for all days of the given time interval to fill them in.

```{r fill in NAs}
activity.noNAs <- activity
for(i in 1:nrow(activity.noNAs)){
if(is.na(activity.noNAs$steps[i])){
    activity.noNAs$steps[i] = byint$steps[which(byint$intervals == 
                                                    activity.noNAs$interval[i])]
}}
```

Excellent, now that all our NAs are filled in, we're ready to graph a new 
histogram!

```{r new histogram}
total.noNAs <- tapply(activity.noNAs$steps, activity.noNAs$date, sum)
hist(total.noNAs)   
```

Interesting...
Let's take a look at the mean and median.

```{r mean median no NAs}
mean(total.noNAs)
median(total.noNAs)
```

Well now, that's interesting! Our histogram has a distribution that more closely
resembles a bell curve! Also, by imputing the NAs we have increased both our 
mean and median. As a matter of fact, the median and the mean are now the same
number! 

What does that tell us? I'd say that the majority of the NAs were from times 
when one could expect to have walked more!

## Are there differences in activity patterns between weekdays and weekends?

Our final task is to look at weekdays versus weekends.

Let's add a factor to denote which days are weekends and which are weekdays.

```{r daytype}
for(i in 1:nrow(activity.noNAs)){
    activity.noNAs$daytype[i] <- 
        ifelse(weekdays(as.Date(activity.noNAs$date[i])) 
               %in% c("Saturday","Sunday"), "Weekend","Weekday")
}
```

Now we want to look at plots of average steps per day with respect to weekends
and weekdays.

First we'll split up our weekend averages and our weekday averages and put them
into a data frame with our intervals

```{r daytype dataframe}
byday <- tapply(activity.noNAs$steps, list(activity.noNAs$interval, 
            activity.noNAs$daytype), mean)
byday <- as.data.frame(byday)
byday$interval <- unique(activity.noNAs$interval)
final <- append(byday$Weekend, byday$Weekday)
final <- as.data.frame(final)
final$intervals <- append(byday$interval, byday$interval)
colnames(final) <- c("Steps","Intervals")
for(i in 1:nrow(final)){
    final$DayType[i] <- ifelse(i <= 288, "Weekend", "Weekday")
}
```

Now we can easily graph our results.

```{r final graph}
library(lattice)
xyplot(Steps ~ Intervals | as.factor(DayType), data = final, type = "l", 
       layout = c(1,2), ylab = "Number of Steps")
```

As we can see, more steps are taken during Weekday mornings, whereas on Weekends
the number of steps taken throughout the day is more varied.